---
title: "Analysing the UP Government Budget"
author: "Hrishant Singhal"
date: "Fiscal Data Fellow Assignment"
output:
  html_document: 
    code_folding: hide
    anchor_sections: FALSE
    toc: yes
    df_print: tibble
  pdf_document: default
  word_document: default
---

The Uttar Pradesh Government budgest data files that have been used in this analysis have been downloaded from [here](https://drive.google.com/drive/folders/1leLy55YFop0PphhcrvRMCirI3gVxJ2ky). 

The census data files for population and the R markdown file on which this code has been written can be downloaded from [here](https://github.com/hrishantsinghal/CivicDataLab-Assignment-).

## 0. Introduction and Primary Data Cleaning while loading the CSV files 

The column heads of 2017, 2018 and 2019 csv files don't match with the variables in the metadata. We have been given "Total Expenditure Upto November" in the csv files instead of "Total Expenditure Upto March". In these files we have not been given any variable to help us know the expenditure in that financial year. 

I assume that it might be a typo so I change the column names to match with the metadata.

```{r message=FALSE, warning=FALSE}
knitr::opts_chunk$set(comment = NA)

library(DT)
library(tidylog)
library(readr)
library(fuzzyjoin)
library(readxl)
library(tidyverse)
library(R.utils)

#Loading the files and adding the year column
up_17<-read.csv("D:/CivicDataLab/district_level_mapping_2017.csv",encoding="UTF-8") %>% 
  mutate(Grant.Number=as.character(Grant.Number)) %>% mutate(year=2017)
up_18<-read.csv("D:/CivicDataLab/district_level_mapping_2018.csv",encoding="UTF-8") %>% mutate(year=2018) %>% 
  mutate(Grant.Number=as.character(Grant.Number))
up_19<-read.csv("D:/CivicDataLab/district_level_mapping_2019.csv",encoding="UTF-8") %>% mutate(year=2019) %>% 
  mutate(Grant.Number=as.character(Grant.Number))
up_20<-read.csv("D:/CivicDataLab/district_level_mapping_2020.csv",encoding="UTF-8") %>% mutate(year=2020) %>% 
  mutate(Grant.Number=as.character(Grant.Number))

#Matching all the column names
names(up_17)<-names(up_20)
names(up_18)<-names(up_20)
names(up_19)<-names(up_20)

#Combining all the files
up<-dplyr::bind_rows(up_17,up_18,up_19,up_20) 

```
## 1. Share of Expenditure on School Education incurred by various Departments/Ministries.

According to the LMMHA, the Major Head Codes 2202 and 4202 represent revenue expenditure and capital expenditure respectively on general education. 

#### The general education head includes all the following:
1. Elementary Education
2. Secondary Education
3. University and Higher Education
4. Adult Education
5. Language Development
6. General

Since the task requires me to only look at school education, I include the expenditures of elementary and secondary education. 

#### Note on data cleaning of the departments/ministries.

The original data has the column "Grant.Head" which does not neatly distinguish all the departments, some departments have been repeated twice or thrice with different suffixes in brackets. I remove all the suffixes and combine the departments together so that every row shows a unique department.

```{r cars, message=FALSE, warning=FALSE}
#Filtering education expenditure
up_education<-up %>% filter(str_detect(Scheme.Code, "^220201|^220202|^420201201|^420201202")) %>% 
  separate(col=`Grant.Head.Description`,sep="[(]",into=c("Departments/Ministries","waste")) %>% select(-waste)

#Removing the spaces so that there is no problem in grouping 
up_education$`Departments/Ministries`<-str_trim(up_education$`Departments/Ministries`)

#Summarizing for each department
up_education_depts<-up_education %>% group_by(`Departments/Ministries`) %>% 
  summarise(`Total Expenditure`=sum(Total.Expenditure.Upto.Month..March.)) %>% 
  group_by(`Departments/Ministries`) %>% summarise(across(where(is.numeric), sum)) %>% 
  mutate(`Share of Total Expenditure`=round(`Total Expenditure`*100/sum(`Total Expenditure`),2))

datatable(up_education_depts)
```

## 2. Share of Capital Expenditure

#### Data Cleaning Process

I take Capital and Revenue expenditures of all the departments. Then I convert this data into a wide format data and mutate the column which shows the share of capital expenditure in the total expenditure. 

```{r pressure, echo=FALSE, message=FALSE, warning=FALSE}
#Grouping by department and type of expenditure
up_education_depts_wide<-up_education %>% group_by(`Departments/Ministries`,Major.Head.Code) %>% 
  summarise(`Total Expenditure`=sum(Total.Expenditure.Upto.Month..March.)) 

#Converting to wide format
cap_rev<-spread(up_education_depts_wide, key = "Major.Head.Code", value = `Total Expenditure`) %>% 
  select(revenue_expenditure="2202",capital_expenditure="4202") %>% 
  mutate(revenue_expenditure=round(revenue_expenditure,2),capital_expenditure=round(capital_expenditure,2))

#Adding zeroes where there is NA
cap_rev[is.na(cap_rev)] = 0

#Estimating the share of capital expenditure
cap_rev<-cap_rev %>% mutate(`Share of Capital Expenditure in Total Expenditure (in percentage)`=capital_expenditure*100/(revenue_expenditure+capital_expenditure)) %>% 
  mutate_if(is.numeric, round,4) %>% 
  rename(`Capital Expenditure`=capital_expenditure,
         `Revenue Expenditure`=revenue_expenditure)

datatable(cap_rev)

```

## 3. Per-Capita Expenditure on School Education based on Projected Population in:
### A. Each District 

#### Working on Projected Population 

I merge the Census 2011 and 2001 Population Datasets. I assume that the population growth rate would remain the same after 2011 as it has remained between 2001 and 2011. Using this assumption, I calculate each of Uttar Pradesh's district's population in 2021. 

I then do linear interpolation to get population estimates for all years between 2011 and 2021. At last I filter the dataset to only contain the years 2017, 2018, 2019 and 2020.

```{r message=FALSE, warning=FALSE}
#Importing Population 2001 and 2011 and doing linear interpolation 
district_data_1654156775 <- read_csv("C:/Users/hrish/Downloads/district_data_1654156775.csv")
population<-district_data_1654156775 %>% filter(statename=="Uttar Pradesh")

#Taking the required columns
population<-population[,c(4,5,7)]

#Changing Column Names
names(population)[names(population)=="districtname"]<-"district"
names(population)[names(population)=="statename"]<-"stateut"
names(population)[names(population)=="value"]<-"pop"

#Using lowercase for districts
population$district<-tolower(as.character(population$district))

#Combining the state and the district columns
population<-population %>% mutate(state_district=paste(stateut,district,sep="!")) %>% 
  relocate(state_district) %>% select(-stateut,-district) %>% mutate(year=2011)

#Importing 2001 population data and doing the required cleaning  
pop_2001<-read.csv("D:/CEDA Internship Work/Harmonising 30/2014,15,16/district_data_1654765263.csv")
pop_2001<-pop_2001 %>% filter(statename=="Uttar Pradesh")

#Taking the required columns
pop_2001<-pop_2001[,c(4,5,7)]

#Changing Column Names
names(pop_2001)[names(pop_2001)=="districtname"]<-"district"
names(pop_2001)[names(pop_2001)=="statename"]<-"stateut"
names(pop_2001)[names(pop_2001)=="value"]<-"pop"

#Using lowercase for districts
pop_2001$district<-tolower(as.character(pop_2001$district))
pop_2001$pop<-round(as.numeric(pop_2001$pop))

#Combining the state and the district columns
pop_2001<-pop_2001 %>% mutate(state_district=paste(stateut,district,sep="!")) %>% 
  relocate(state_district) %>% select(-stateut,-district) %>% mutate(year=2001)

#Combining 2001 and 2011 populations
population<-population %>% bind_rows(pop_2001) %>% arrange(state_district,year)

#taking only state_districts which are there in both census
d_in_both<-population %>% group_by(state_district) %>% summarise(n=n()) %>% filter(n>=2) %>% 
  pull(state_district)
population<-population %>% filter(state_district %in% d_in_both)

#Population 2021
pop_2011_p<-population %>% filter(year==2011) %>% mutate(pop=2*pop)  
pop_2001_p<-population %>% filter(year==2001)
pop_2021<-data.frame(state_district=pop_2011_p$state_district,pop=pop_2011_p$pop-pop_2001_p$pop,year=2021)
population<-population %>% bind_rows(pop_2021) %>% arrange(state_district,year) %>% filter(year!=2001)

#Interpolating 2011 and 2021 populations
interpolate_func <- function(variable, data) {
  data %>% 
    group_by(state_district) %>% 
    # can't interpolate if only one year
    filter(n() >= 2) %>% 
    group_modify(~as_tibble(approx(.x$year, .x[[variable]], 
                                   xout = min(.x$year):max(.x$year)))) %>% 
    set_names("state_district", "year", paste0(variable, "_interpolated")) %>% 
    ungroup()
}

vars_to_interpolate <- names(select(population, -state_district, -year))

population<-map(vars_to_interpolate, interpolate_func, 
                data = population) %>% 
  reduce(full_join, by = c("state_district", "year")) 

#Taking the population of the required years 
population<-population %>% rename(pop=pop_interpolated) %>% mutate(pop=round(pop)) %>% 
  separate(col=state_district,sep="!",into=c("state","district")) %>% filter(year %in% c(2017,2018,2019,2020))
population$district<-capitalize(population$district)
population_capital<- population %>% 
  rename(Population=pop,
         State=state,
         District=district,
         Year=year)
datatable(population_capital)

population$district<-tolower(as.character(population$district))
```

#### Direct Merge Attempt 
The population census data has 71 districts in Uttar Pradesh whereas there are 80 treasuries. We try to merge the UP budget data with the population data after doing some basic data cleaning like renaming 'Prayagraj' to 'Allahbad' and renaming 'lucknow coll.' to 'lucknow' and 'lucknow-jb' to 'lucknow'. 

```{r message=FALSE, warning=FALSE}
up_education$Treasury<-tolower(up_education$Treasury)
population$district<-tolower(population$district)
up_education$Treasury[up_education$Treasury=="prayagraj-coll"]<-"allahabad"
up_education$Treasury[up_education$Treasury=="allahabad ii"]<-"allahabad"
up_education$Treasury[up_education$Treasury=="prayagraj-civ"]<-"allahabad"
up_education$Treasury[up_education$Treasury=="lucknow coll."]<-"lucknow"
up_education$Treasury[up_education$Treasury=="lucknow-jb"]<-"lucknow"

up_education_districts<-up_education %>% group_by(Treasury,year)%>% 
  summarise(`Total Expenditure`=sum(Total.Expenditure.Upto.Month..March.)) %>% rename(district=Treasury)

merge_attempt<-population %>% inner_join(up_education_districts,by="district") %>% filter(year.x==2020,year.y==2020)

datatable(merge_attempt)

```
We see that a direct merge has only 49 districts. Since the direct merge attempt leaves out 22 districts, we try to use a package called fuzzywuzzy to match districts that have been left out so that districts that have a very small difference in their spellings can be merged. 

#### Fuzzywuzzy Attempt 

```{r message=FALSE, warning=FALSE}
districts_removed<-merge_attempt$district

fuzzy_attempt<-stringdist_join(up_education_districts %>% filter(!district %in%districts_removed), 
                               population %>% filter(!district %in%districts_removed),
                               by = c("district"),
                               mode = "full",
                               ignore_case = TRUE, 
                               max_dist = 1,
                               method="jw",
                               distance_col = "dist") %>%
  group_by(district.x) %>% filter(year.x==year.y) %>% 
  slice_min(order_by = dist, n = 1)
fuzzy_attempt_display<-fuzzy_attempt %>% filter(year.x==2020) %>% mutate(dist=round(dist,2)) %>% select(-year.y) %>% relocate(district.x,district.y)

datatable(fuzzy_attempt_display)

```

Though we see some benefits of following this approach with districts like 'ambedkarnagar' getting matched to 'ambedkar nagar' and 'badaun' getting matched to 'budaun', there are a lot of errors here too like 'ayodhya' being matched to 'baghpat' and 'bhadohi' getting matched to 'jhansi'. 

Such things are hard to be matched using only code, for example, Ayodhya District is also knows as Faizabad District. We will have to rename it manually since the census calls it Faizabad. Similarly, a lot of such places will have to be renamed manually after doing some research.

#### Combining Fuzzywuzzy with Manual Work 

I google the districts that could not be matched and replace their names with their other names or something they are a part of. For example, Shamli is a part of Saharanpur which is there in the 2011 census data, Hathras is a part of Mahamaya Nagar, GB Nagar is actually called Gautam Buddha Nagar and other such editing was done. 

The only subjective judgment that has been done while matching was that the treasury "sadar try-hmrpr" was matched to "hamirpur". 

Hapur could not be matched to any district from the population census. 

In the following table, the names of all districts have been used according to their census 2011 names.  
```{r message=FALSE, warning=FALSE}
#After googling and changing the names
up_education$Treasury[up_education$Treasury=="ayodhya"]<-"faizabad"
up_education$Treasury[up_education$Treasury=="g.b.nagar"]<-"gautam buddha nagar"
up_education$Treasury[up_education$Treasury=="j.p.nagar"]<-"jyotiba phule nagar"
up_education$Treasury[up_education$Treasury=="sambhal"]<-"moradabad"
up_education$Treasury[up_education$Treasury=="shamli"]<-"saharanpur"
up_education$Treasury[up_education$Treasury=="orai"]<-"jalaun"
up_education$Treasury[up_education$Treasury=="csmaharaj nagar"]<-"sitapur"
up_education$Treasury[up_education$Treasury=="hathras"]<-"mahamaya nagar"
up_education$Treasury[up_education$Treasury=="bhadohi"]<-"sant ravidas nagar (bhadohi)"
up_education$Treasury[up_education$Treasury=="sadar try-hmrpr"]<-"hamirpur"

up_education_districts<-up_education %>% group_by(Treasury,year)%>% 
  summarise(`Total Expenditure`=sum(Total.Expenditure.Upto.Month..March.)) %>% rename(district=Treasury)

#Trying fuzzy
fuzzy_attempt<-stringdist_join(up_education_districts, population,
                               by = c("district"),
                               mode = "full",
                               ignore_case = TRUE, 
                               max_dist = 2,
                               method="jw",
                               distance_col = "dist") %>%
  group_by(district.x) %>% filter(year.x==year.y) %>% 
  slice_min(order_by = dist, n = 1)

#The districts 'badaun' was matched to 2 districts. The mismatch has been left out.
fuzzy_attempt<-fuzzy_attempt[-c(25, 27, 29,31),]

#Doing some editing in the dataframe to make it presentable.
per_capita_district<-fuzzy_attempt %>% ungroup() %>% select(-district.x,-dist,-year.y) %>% 
  rename(District=district.y,Year=year.x,Population=pop) %>% 
  relocate(District) %>%
  mutate(`Per Capita Expenditure`=round(as.numeric(`Total Expenditure`)/as.numeric(Population),2)) %>% select(-state) %>% mutate(District=capitalize(District))

datatable(per_capita_district)

```

All the 71 districts in population have been matched to the treasuries column in the budget data.

### B. The State
Lucknow's expenditure on school education is very high and that increases the average expenditure of the state on education.

```{r message=FALSE, warning=FALSE}
per_capita_state<-per_capita_district %>% mutate(State="Uttar Pradesh") %>% 
  group_by(State,Year) %>% 
  summarise(`Per Capita Expenditure`=round(sum(`Total Expenditure`)/sum(`Population`),2))

datatable(per_capita_state)
```


## 4. Districts Ranked based on Utilization of Allotted Funds

For calculating the utilization of allotted funds, we take the sum of the total expenditure and divide it by the sum of progressive allotment. 

All districts have been ranked in decreasing order based on their utilization of allotted funds.

### A. Revenue Expenditure 

```{r}
new_names<-data.frame(Treasury=fuzzy_attempt$district.x,District=fuzzy_attempt$district.y)

up_education_allocation_revenue<-up_education %>% inner_join(new_names,by="Treasury") %>% filter(Major.Head.Code==2202) %>% group_by(District) %>% 
  summarise(`Utilisation of Funds`=round(100*sum(Total.Expenditure.Upto.Month..March.)/sum(Progressive.Allotment),2)) %>% 
  arrange(-`Utilisation of Funds`)

#Capitalizing the first letter of the district's names.
up_education_allocation_revenue$District<-capitalize(up_education_allocation_revenue$District)
datatable(up_education_allocation_revenue)

```

### B. Capital Expenditure
```{r}
up_education_allocation_capital<-up_education %>% inner_join(new_names,by="Treasury") %>% filter(Major.Head.Code==4202) %>% group_by(District) %>% 
  summarise(`Utilisation of Funds`=round(100*sum(Total.Expenditure.Upto.Month..March.)/sum(Progressive.Allotment),2)) %>% 
  arrange(-`Utilisation of Funds`)

#Capitalizing the first letter of the district's names.
up_education_allocation_capital$District<-capitalize(up_education_allocation_capital$District)
datatable(up_education_allocation_capital)
```






